### 2021-11-16 publication/experiments.tex
Final IDA draft

\section{Experimental Results}
\label{sec:experiments}

@@

\begin{itemize}
	\item \textbf{Political books} \cite{polbooks} ($N=105, E=441, D=3$) -- network of Amazon political books published close to the 2004 presidential election. Two books are connected if they were frequently co-purchased. Vertex features encode the political affiliation of the author (liberal, conservative, or neutral).
	\item \textbf{Primary school dynamic contacts} \cite{schools} ($N=238, E=5539, D=13$) -- network of 238 individuals (students and teachers),
with edges denoting face-to-face contacts at a primary school in Lyon, France. 
The vertex features are class membership (one of 10 values: 1A-5B), gender (male, female), and status (teacher, student). We choose to analyse just the second day of results.
	\item \textbf{Facebook egonet} \cite{fb-snap} ($N=747, E=30025, D=480$) -- network of Facebook users with edges denoting ``friends''.
Vertex features are fully anonymised and encode information about each user's education history, languages spoken, gender, home-town, birthday etc. We focus on the egonet with id 1912.
\end{itemize}
%
For reference, the inferred partitions for all of these are given on Figure~\ref{fig:graphs-all}.
We employ the following metrics to assess model performance. 
First, the average
description length per entity (nodes and edges) 
$\bar{S}_e$ 
used to gauge the SBM fit is defined as:
%
\begin{equation}
	\bar{S}_e \coloneqq \frac{1}{(N+E) |\Tcal_b|} \sum_{t\in \Tcal_b} S \left( b^{(t)} \right).
	\label{eqn:mean-dl}
\end{equation}
%
Next, to assess the performance of the feature-to-block predictor, 
the vertex set $[N]$ 
is partitioned at random so that 
a constant fraction $f$ of vertices form the training set $\Gcal_0$ and 
the remainder form the test set $\Gcal_1$.
The $b$-chain is run using the whole network but we $\in \Gcal_0$ is 
used to train the $\theta$-chain. 
% As $|\Gcal_0| \neq |\Gcal_1|$, in general, 
Then the  the average cross-entropy loss 
over each set is used to gauge the quality of the fit,
%
\begin{equation}
	\bar{\Lcal}_\star \coloneqq \frac{1}{|\Tcal_\theta|} \sum_{t \in \Tcal_\theta} \Lcal_\star^{(t)},
	\quad \textrm{where} \quad
	\Lcal_\star^{(t)} \coloneqq \frac{1}{|\Gcal_\star|} \sum_{i \in \Gcal_\star}\sum_{j \in [B]} \hat{y}_{ij} \log \frac{1}{\phi_j \left(x_i; \theta^{(t)} \right)},
	\label{eqn:cross-entropy-loss}
\end{equation}
%
where $\star \in \{0, 1\}$ toggles between the training and test sets.
and $\hat{y}_{ij}$ is defined in~(\ref{eqn:y-hat}).
%
Nevertheless, the cross-entropy loss is a coarse measure of fit. 
A new measure, specific to each detected block,
can be defined as follows. Let
$\Bcal_\star(j)$ 
be the set of vertices with maximum a posteriori probability of belonging 
to block~$j$,
$
	\Bcal_\star(j) \coloneqq \{i \in \Gcal_\star : \hat{b}_i = j\},
$
where
$ 
	\hat{b}_i \coloneqq \argmax_j \hat{y}_{ij},
$
and
define the {\em block-accuracy} for block $j$ as,
%
\begin{equation}
	\eta_\star(j) \coloneqq \frac{1}{|\Bcal_\star (j)| \cdot 
	|\Tcal_\theta| } 
	\sum_{i \in \Bcal_\star (j)}  \sum_{t \in \Tcal_\theta}
	\one \left\{\hat{b}_i = \argmax_j \phi_j \left( x_i; \theta^{(t)} \right) \right\}.
	\label{eqn:accuracy}
\end{equation}
%
This effectively tests whether the feature-to-block and 
graph-to-block predictions agree in their largest component.
For the higher-dimensional datasets, we also apply the 
dimensionality reduction method 
of Section~\ref{sec:dim-reduction}.  
We then retrain the feature-block predictor using only the retained 
feature set $\Dcal'$, and report the log-loss over the training and 
test sets for the reduced classifier -- 
denoted $\bar{\Lcal}_0'$ and $\bar{\Lcal}_1'$ respectively. 
\input{experiments-figures}
\input{experiments-detailed}

>>>>

\section{Experimental results}
\label{sec:experiments}

@@

\begin{itemize}
	\item \textbf{Political books} \cite{polbooks} ($N=105, E=441, D=3$) -- network of Amazon political books published close to the 2004 presidential election. Two books are connected if they were frequently co-purchased. Vertex features encode the political affiliation of the author (liberal, conservative, or neutral).
	\item \textbf{Primary school dynamic contacts} \cite{schools} ($N=238, E=5539, D=13$) -- network of 238 individuals (students and teachers),
with edges denoting face-to-face contacts at a primary school in Lyon, France. 
The vertex features are class membership (one of 10 values: 1A-5B), gender (male, female), and status (teacher, student). We choose to analyse just the second day of results.
	\item \textbf{Facebook egonet} \cite{fb-snap} ($N=747, E=30025, D=480$) -- network of Facebook users with edges denoting ``friends''.
Vertex features are fully anonymised and encode information about each user's education history, languages spoken, gender, home-town, birthday etc. We focus on the egonet with id 1912.
\end{itemize}
%
For reference, the inferred partitions for all of these are given on Figure~\ref{fig:graphs-all}.
We employ the following metrics to assess model performance. 
First, the average
description length per entity (nodes and edges) 
$\bar{S}_e$ 
used to gauge the SBM fit is defined as:
%
\begin{equation}
	\bar{S}_e \coloneqq \frac{1}{(N+E) |\Tcal_b|} \sum_{t\in \Tcal_b} S \left( b^{(t)} \right).
	\label{eqn:mean-dl}
\end{equation}
%
Next, to assess the performance of the feature-to-block predictor, 
the vertex set $[N]$ 
is partitioned at random so that 
a constant fraction $f$ of vertices form the training set $\Gcal_0$ and 
the remainder form the test set $\Gcal_1$.
The $b$-chain is run using the whole network but only vertices $v \in \Gcal_0$ are 
used for the $\theta$-chain. 
Then the  the average cross-entropy loss 
over each set is used to gauge the quality of the fit,
%
\begin{equation}
	\bar{\Lcal}_\star \coloneqq \frac{1}{|\Tcal_\theta|} \sum_{t \in \Tcal_\theta} \Lcal_\star^{(t)},
	\quad \textrm{where} \quad
	\Lcal_\star^{(t)} \coloneqq \frac{1}{|\Gcal_\star|} \sum_{i \in \Gcal_\star}\sum_{j \in [B]} \hat{y}_{ij} \log \frac{1}{\phi_j \left(x_i; \theta^{(t)} \right)},
	\label{eqn:cross-entropy-loss}
\end{equation}
%
where $\star \in \{0, 1\}$ toggles between the training and test sets
and $\hat{y}_{ij}$ is defined in~(\ref{eqn:y-hat}).
%
Nevertheless, the cross-entropy loss is a coarse measure of fit. 
A new measure, specific to each detected block,
can be defined as follows. Let
$\Bcal_\star(j)$ 
be the set of vertices with maximum a posteriori probability of belonging 
to block~$j$,
$
	\Bcal_\star(j) \coloneqq \{i \in \Gcal_\star : \hat{b}_i = j\},
$
where
$ 
	\hat{b}_i \coloneqq \argmax_j \hat{y}_{ij},
$
and
define the {\em block-accuracy} for block $j$ as,
%
\begin{equation}
	\eta_\star(j) \coloneqq \frac{1}{|\Bcal_\star (j)| \cdot 
	|\Tcal_\theta| } 
	\sum_{i \in \Bcal_\star (j)}  \sum_{t \in \Tcal_\theta}
	\one \left\{\hat{b}_i = \argmax_j \phi_j \left( x_i; \theta^{(t)} \right) \right\}.
	\label{eqn:accuracy}
\end{equation}
%
This effectively tests whether the feature-to-block and 
graph-to-block predictions agree in their largest component.
For the higher-dimensional datasets, we also apply the 
dimensionality reduction method 
of Section~\ref{sec:dim-reduction}.  
We then retrain the feature-block predictor using only the retained 
feature set $\Dcal'$, and report the log-loss over the training and 
test sets for the reduced classifier -- 
denoted $\bar{\Lcal}_0'$ and $\bar{\Lcal}_1'$ respectively. 
\input{experiments-figures}
\input{experiments-detailed}

### 2021-11-16 publication/ffbm.tex
Final IDA draft

In this section we propose a novel generative model for labelled networks. We call this the feature-first block model (FFBM),
illustrated in Figure~\ref{fig:ffbm}).

Let $N$ denote the number of vertices, $B$ the number of blocks
and ${\cal X}$ the set of values each feature can take.
We define the vector $x_i \in \Xcal^D$ as the feature vector for vertex $i$, 
where $D$ is the number of features associated with each vertex.
For example, in the datasets we analyse we deal with binary feature flags
(denoting the presence/absence of each feature),
so $\Xcal = \{0, 1\}$. We write $X$ for the $N\times D$ {\em feature matrix} containing
the feature vectors $\{x_i\}_{i=1}^{N}$ 
as its rows.

@@

To complete the description of our Bayesian framework,
priors on $\theta$ and $\psi$ must also be specified. 
We place a Gaussian prior on $\theta$ such that
each element of $\theta$ has an independent ${\cal N}(0,\sigma_\theta^2)$
prior, with hyperparameter $\sigma_\theta^2$:
%
\begin{equation}
	p(\theta) \sim \Gaussian \left( \theta ; 0, \sigma_\theta^2 I \right).
	\label{eqn:theta-prior}
\end{equation}
%
This choice of prior gives a very
simple form for the conditional distribution of the block membership vector $b$ given $X$; it is a uniform distribution:
%
\begin{equation}
	p(b | X) = \int p(b | X, \theta) p(\theta) d\theta = B^{-N}.
	\label{eqn:b-pseudo-prior}
\end{equation}
%
The proof is given in Appendix~\ref{appdx:b|x}. This an important simplification as evaluating $p(b | X)$ does not require an expensive  integration over $\theta$ nor does it depend on $X$.
Peixoto \cite{Peixoto-Bayesian-Microcanonical} proposes careful choices for 
the priors on the additional microcanonical SBM parameters $\psi$, which we adopt without repeating their exact form here. 
The idea is to write the joint distribution on $(b, e, k)$ as a product of 
conditionals, $p(b, e, k) = p(b) p(e | b) p(k | e, b)= p(b) p(\psi | b)$. 
In our case, conditioning on $X$ is also necessary, leading to,
$
	p(b, \psi | X) = p(b | X) p(\psi | b, X) = p(b | X) p(\psi | b),
$
where we used the fact $\psi$ and $X$ are conditionally 
independent given $b$.
All that concerns the main argument is that $p(\psi|b)$ has
an easily computable form.

>>>>

In this section we propose a novel generative model for labelled networks. We call this the feature-first block model (FFBM),
illustrated in Figure~\ref{fig:ffbm}.

Let $N$ denote the number of vertices, $B$ the number of blocks
and ${\cal X}$ the set of values each feature can take.
We define the vector $x_i \in \Xcal^D$ as the feature vector for vertex $i$, 
where $D$ is the number of features associated with each vertex.
For example, in the datasets we analyse, we deal with binary feature flags
(denoting the presence/absence of each feature),
so $\Xcal = \{0, 1\}$. We write $X$ for the $N\times D$ {\em feature matrix} containing
the feature vectors $\{x_i\}_{i=1}^{N}$ 
as its rows.

@@

To complete the description of our Bayesian framework,
priors on $\theta$ and $\psi$ must also be specified. 
We place a Gaussian prior on $\theta$ such that
each element of $\theta$ has an independent ${\cal N}(0,\sigma_\theta^2)$
prior, with hyperparameter $\sigma_\theta^2$:
%
\begin{equation}
	p(\theta) \sim \Gaussian \left( \theta ; 0, \sigma_\theta^2 I \right).
	\label{eqn:theta-prior}
\end{equation}
%
This choice of prior gives a very
simple form for the conditional distribution of the block membership vector $b$ given $X$; it is a uniform distribution:
%
\begin{equation}
	p(b | X) = \int p(b | X, \theta) p(\theta) d\theta = B^{-N}.
	\label{eqn:b-pseudo-prior}
\end{equation}
%
The proof is given in Appendix~\ref{appdx:b|x}. This is an important simplification as evaluating $p(b | X)$ does not require an expensive  integration over $\theta$ nor does it depend on $X$.
Peixoto \cite{Peixoto-Bayesian-Microcanonical} proposes careful choices for 
the priors on the additional microcanonical SBM parameters $\psi$, which we adopt without repeating their exact form here. 
The idea is to write the joint distribution on $(b, e, k)$ as a product of 
conditionals, $p(b, e, k) = p(b) p(e | b) p(k | e, b)= p(b) p(\psi | b)$. 
In our case, conditioning on $X$ is also necessary, leading to,
$
	p(b, \psi | X) = p(b | X) p(\psi | b, X) = p(b | X) p(\psi | b),
$
where we used the fact $\psi$ and $X$ are conditionally 
independent given $b$.
All that concerns the main argument is that $p(\psi|b)$ has
an easily computable form.

### 2021-11-16 publication/prelim.tex
Final IDA draft

We first need a model for community-like structure in a network. For this we adopt the widely-used stochastic block model (SBM). This is a latent variable model where each vertex belongs to a single block and the probability two nodes are connected depends only on the block memberships of each.
Specifically, we will use the microcanonical variant of the SBM, proposed by Peixoto \cite{Peixoto-Bayesian-Microcanonical}. To allow for degree-variability between members of the same block, we adopt the following degree-corrected 
formulation (DC-SBM):
% , defined in~(\ref{defn:microcan-dc-sbm}).

>>>>

We first need a model for community-like structure in a network. For this we adopt the widely-used stochastic block model (SBM). This is a latent variable model where each vertex belongs to a single block and the probability two vertices are connected depends only on the block memberships of each.
Specifically, we will use the microcanonical variant of the SBM, proposed by Peixoto \cite{Peixoto-Bayesian-Microcanonical}. To allow for degree-variability between members of the same block, we adopt the following degree-corrected 
formulation (DC-SBM):
% , defined in~(\ref{defn:microcan-dc-sbm}).
